---
output: 
  html_document: 
    fig_caption: yes
---
"Reproducible Research: Peer Assessment 1" 
============================
## by 'Gopalkriz'  
### Date: 
```{r echo=FALSE}

Sys.Date()
```  
### Coursera - Reproducible Research (Aug 2015 batch)  

## Project Introduction  
### Dataset: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip  
### ProjectFolder: https://github.com/gopalkriz/RepData_PeerAssessment1  
### Local Repository:
```{r ,echo=FALSE}

print("Cloned from github D:/VIVEK/DataAnalysis/Coursera/5 RepRes/RepData_PeerAssessment1")
```  

-------------------

##Setting Global Options  
arguments set for code chunks to be included in output  
```{r setoptions, echo=TRUE}

library(knitr)
opts_chunk$set(echo = TRUE, results = "asis")
```

##Set Working Directory with required Folder Path
```{r Set Working Directory}

WDoriginal <- getwd()
setwd("D:/VIVEK/DataAnalysis/Coursera/5 RepRes/RepData_PeerAssessment1")
```

-------------------

## Project Introduction
About: "quantified self" movement - collect data of personal activity using monitoring devices
Data: collected at 5 minute intervals through out the day for two months from an anonymous individual during October and November, 2012.
Variable: steps taken in 5 minute intervals each day.

Dataset: https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip
ProjectFolder: https://github.com/gopalkriz/RepData_PeerAssessment1
Local Repository:
```{r, echo=FALSE}

print("Cloned from github D:/VIVEK/DataAnalysis/Coursera/5 RepRes/RepData_PeerAssessment1")
``` 

-------------------

## Loading and preprocessing the data
###Upload the Data
```{r Upload the Data}

unzip("activity.zip")
Data <- read.csv("activity.csv", header = TRUE, sep =",")
```

###Understand the Data
dataset: comma-separated-value (CSV) file 17,568 observations
Three variables included in this dataset are:
steps: taking in a 5-minute interval (missing values are coded as NA)
date: on which the measurement was taken in YYYY-MM-DD format
interval: continuous counter Identifier of time for the measurement
```{r Understand the Data}

object.size(Data)
class(Data)
dim(Data); nrow(Data); ncol(Data)
names(Data)
str(Data)
head(Data)
tail(Data)
summary(Data)
```

-------------------

##Question 1.0
### What is mean total number of steps taken per day?
note: you can ignore the missing values in the dataset

1.1. Calculate the total number of steps taken per day
### Rcode 1.1 
```{r total steps per day}

daysteps <- aggregate(steps ~ date, data = Data, FUN = sum)
```

1.2. Plot the total number of steps taken per day

Barplot steps-datewise
### Rcode 1.2a
```{r Barplot_DaySteps Plot_001}

par(mfrow = c(1, 1))
par(mar=c(4,4,1,0.5))

plot_001 <- barplot(daysteps$steps, names.arg = daysteps$date, xlab = "date", ylab = "steps", main = "Activity: total steps per day")

dev.copy(png,'plot_001.png')
invisible(plot_001)
dev.off()
```

Histogram steps per day
### Rcode 1.2b
```{r Hist_DaySteps Plot_002}

par(mfrow = c(1, 1))
par(mar=c(4,4,1,0.5))

plot_002 <- hist(daysteps$steps, xlab = "steps", ylab = "days counts", main ="Daywise Activity Histogram") # Histogram of steps per day

dev.copy(png,'plot_002.png')
invisible(plot_002)
dev.off()
```

1.3. Calculate and report the mean and median of the total number of steps taken per day
### Rcode 1.3
```{r mean total steps per day}

mean(daysteps$steps) #Ans: should be 10766.19
median(daysteps$steps) #Ans:  should be 10765
```

##Answer Mean total number of steps taken per day is as per output above.

-------------------
##Question 2.0
### What is the average daily activity pattern?

##Answer 2.0 
```{r average daily actvity}
dayAvg <- aggregate(steps ~ date, data = Data, FUN = mean)
head(dayAvg)
plot(dayAvg)
```

2.1 Make a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis)

### Rcode 2.1
```{r steps per interval}

StepsGroup <- aggregate(steps ~ interval, data = Data, FUN = mean)
```

```{r timeseries Plot_003}

par(mfrow = c(1, 1))
par(mar=c(4,4,1,0.5))

plot_003 <- plot(StepsGroup, type = "l", xlab = " Interval", ylab= "Average of steps", main= "Timeseries steps avg by interval")

dev.copy(png,'plot_003.png')
invisible(plot_003)
dev.off()
```

2.2 Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?
### Rcode 2.2
```{r interval with max step}

StepsGroup$interval[which.max(StepsGroup$steps)] #Ans: 835th Interval
print("with the value as average:")
StepsGroup$steps[which.max(StepsGroup$steps)] #Ans: 206.1698
```
##Answer 2.2
As per above output.

-------------------

##Question 3.0
### Imputing missing values

3.1 Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with NAs)
### Rcode 3.1
```{r find missing values}

sum(is.na(Data)) #Ans: 2304 NAs in entire dataset
sum(is.na(Data$steps)) #col1 has all the NAs
sum(is.na(Data$date)) #col2 does not have missing values
sum(is.na(Data$interval)) #col3 does not have missing values
```
##Answer 3.1
There are 2304 NAs in the data set, all of them in the column with steps measurements

3.2 Devise a strategy for filling in all of the missing values in the dataset.
##Answer 3.2
Method used: means of 5-min intervals as fillers for missing values.
Existing values will remain. NA values will be replaced by values from
StepsGroup <- aggregate(steps ~ interval, data = Data, FUN = mean)

3.3 Create a new dataset that is equal to the original dataset but with the missing data filled in.
### Rcode 3.3
```{r replace NA - clean dataset}
library(Hmisc)

cleandata_mean <- Data
cleandata_mean$steps <- impute(Data$steps, fun=mean)
barplot(as.numeric(cleandata_mean$steps))

cleandata_median <- Data
cleandata_median$steps <- impute(Data$steps, fun=median)
barplot(as.numeric(cleandata_median$steps))
```

3.4a Make a histogram of the total number of steps taken each day.
### Rcode 3.4a
```{r daily steps from cleandata}
dailysteps_mean <- aggregate(steps ~ date, data = cleandata_mean, FUN = sum)
dailysteps_median <- aggregate(steps ~ date, data = cleandata_median, FUN = sum)
```

```{r Hist_DaySteps plot_004}

par(mfrow = c(1, 2))
par(mar=c(4,4,1,0.5))

plot_004a <- hist(dailysteps_mean$steps, main = "histogram - cleandata_mean") # Histogram of steps per day (clean data)

plot_004b <- barplot(dailysteps_mean$steps, names.arg = dailysteps_mean$date, xlab = "date", ylab = "steps", col="light blue", main = "barplot: cleandata_mean")

dev.copy(png,'plot_004.png')
dev.off()
```

3.4b Calculate and report the mean and median total number of steps taken per day.
### Rcode 3.4b
```{r mean of daily steps: cleaned dataset}

#Ans: This was 10766.19  from data with NA  (ie from daysteps$steps)
mean(dailysteps_mean$steps)
#10766.19 mean replace NAs (no difference)
mean(dailysteps_median$steps)
#9354.23 median replace NAs

#Ans: This was 10765 from data with NA (ie from daysteps$steps)
median(dailysteps_mean$steps)
#10766.19 mean replace NAs (minimal difference)
median(dailysteps_median$steps)
#10395 median replace NAs
```

###Selection of Mean as impute. Cleaned dataset is used for further analysis
Data <- cleandata_mean

3.4c Do these values differ from the estimates from the first part of the assignment?
##Answer 3.4c
When the impute is Mean is no / minimal difference
When the impute is Median there is significant difference
Hence, for this dataset, the choice of mean is a better option for impute.

3.4d What is the impact of imputing missing data on the estimates of the total daily number of steps?
##Answer 3.4d
Comparing daysteps (data with NAs) vs dailysteps (cleaned dataset), we do not see much 
diffrence. The impact in minimal for our stratergy of imputing missing data.
-------------------

##Question 4.0
### Are there differences in activity patterns between weekdays and weekends?
note: dataset with the filled-in missing values

4.1 Create a new factor variable in the dataset with two levels -- "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

Create funtion to know daytype (weekday/weekend)
### Rcode 4.1a
```{r function_daytype}

daytype <- function(date) {
    whichday <- weekdays(date)
    if (whichday %in% c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")) 
        return("weekday")
    else (whichday %in% c("Saturday", "Sunday"))
        return("weekend")
}
```

Apply function_daytype on Data
### Rcode 4.1b
```{r find_weekday/weekend}

Data$date <- as.Date(Data$date)
Data$daytype <- sapply(Data$date, FUN=daytype)

head(Data); table(Data$daytype)
```

4.2 Make a panel plot containing a time series plot (i.e. type = "l") of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

### Rcode 4.2
```{r}
weekdayAvg <- aggregate(steps ~ interval, data = Data, subset = Data$daytype == "weekday", FUN = mean)
weekendAvg <- aggregate(steps ~ interval, data = Data, subset = Data$daytype == "weekend", FUN = mean)
```

Plotting the comparison
```{r timeseries Plot_005}

par(mfrow = c(2, 1))
par(mar=c(3,2,2,0.5))

plot_005a <- plot(weekdayAvg , type = "l", main = "Weekday Activty - avg steps", col = "lightblue", lwd=3)

plot_005b <- plot(weekendAvg, type = "l", main = "Weekend Activty - avg steps", col = "green", lwd=2)

dev.copy(png,'plot005.png')
dev.off()
```

```{r timeseries Plot_006}

par(mfrow = c(1, 1))
par(mar=c(3,2,2,0.5))

plot_006 <- plot(weekdayAvg , type = "l", col = "lightblue", lwd =3, xlab="interval", ylab="avg steps", main ="Activty - avg steps")
lines(weekendAvg, type = "l", col = "green", lwd=2)
legend("topright", legend = "Weekend:green Weekday:blue", box.col=8)

dev.copy(png,'plot006.png')
dev.off()
```

##Answer 4.0
Yes, there is a difference in patters of activity between weekdays and weekends?

Weekday activity sees sudden rise start at interval 500.
There is lot more activity on weekends for interval 1000-2000

-------------------

###reset the workign directory to original
setwd(WDoriginal);getwd()
###clear the workspace
rm(age, age.i, cleandata_mean, cleandata_median, dailysteps, dailysteps_mean, dailysteps_median, Data, dayAvg, daysteps, daytype, MainData, naughts, plot_001, Plot_002, plot_003, Plot_004a, Plot_004b, plot_005a, plot_005b, plot_006, StepsGroup, WDoriginal, weekdayAvg, weekendAvg)
ls()
###end
-------------------